Estimating duration in the presence of misclassification

### 2024-08-01 cis-imperfect-testing.tex
improve figure appearance
Christl p.107 comment

Applying the method of \cref{E-perf-test} to CIS data estimates a median duration of less than 14 days, several days shorter than the estimates from \cref{E-ATACCC} (see \cref{imperf-test:fig:problem-cis-estimates}).
The \cref{E-ATACCC} estimates are more reliable over the first 2--3 weeks due to ATACCC's frequent data collection (see \cref{E-biology-data:sec:ataccc,E-ATACCC:sec:discussion}).
This suggests an issue in the estimation using CIS data.
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/CIS_perfect}
  \caption[Estimating survival using CIS data assuming perfect testing]{Estimates of the survival function using CIS data alongside the ATACCC-based estimates of \cref{E-perf-test}. \label{imperf-test:fig:problem-cis-estimates}}
\end{figure}

@@

% Discussion and interrogation of the data led me to
I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negative results.
There are several strands of evidence supporting this hypothesis.
First, it is well known that RT-PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
Furthermore, in \cref{imperf-test:sec:simulate}, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives.
This reproduced the CIS data more faithfully (see \cref{imperf-test:fig:sim-single-pos}) and exhibited the same issues when estimating the duration distribution (not shown).
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/sim-single-positive-episodes}
  \caption[Single positive episodes in CIS simulation]{%
    Proportion of detected episodes which have a single positive episode.
    y-axis is the number of simulations, out of \numprint{1000}, with the indicated proportion.
    Test sensitivity is $\psens$.
    The CIS data are shown as a vertical line.
  }
  \label{imperf-test:fig:sim-single-pos}
\end{figure}

@@

The qualitative increase in long episodes is robust to the choice of prior for $\ntot$, the assumed value for $\psens$ (see \cref{imperf-test:fig:cis-sensitivity}), and the choice of prior for the hazards, $\lambda_t$ (see \cref{E-perf-test:sec:parameters-priors}).
However, the survival proportion over the first 4 weeks is sensitive to these choices.
The estimate using a test sensitivity of 0.8 and $\NBc(\mu\inform, r\inform)$ give a median survival time most similar to that of \cref{E-ATACCC}.
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/CIS_final}
  \caption[Comparing estimates of the duration distribution.]{
    Duration estimates using the model in this chapter (CIS) alongside those from \cref{E-ATACCC} (ATACCC).
    Uses the informative prior on $\ntot$, the model combination prior for the hazards, and $\psensi = 0.8$.
  }
  \label{imperf-test:fig:cis-estimates}
\end{figure}
\begin{figure}
  \thisfloatpagestyle{empty}
  \makebox[\textwidth][c]{\includegraphics{cis-imperfect-testing/CIS_vary}}
  \caption[Sensitivity of CIS estimates to the prior.]{%
    (A-C) Sensitivity of CIS estimates to the value of $r$, the dispersion of the prior for $\ntot$ (a larger $r$ means a more informative prior) when $\psens = 0.8$.
    (D-F) Sensitivity of CIS estimates to the value of $\psens$ when $r = r\inform$.
    A and D: median survival time, in comparison to the ATACCC-based estimate of \cref{E-ATACCC} (shown in red).
    B and E: survival at day 50, $S_{\vec{\theta}}(50)$.
    C and F: full survival curves out to day 100.
  }
  \label{imperf-test:fig:cis-sensitivity}
  \todo[inline]{Optional improvement: Could highlight the "preferred" model which exists in both rows and is the one used subsequently, and/or include ATACCC in the far-right hand column}
\end{figure}

The estimates are sensitive to the choice of $r$, the strength of the prior on $\ntot$.
A low value for $r$, giving a very weak, almost uninformative, prior on $\ntot$ causes its posterior estimate to be much higher than the estimate from \textcite{birrellRTM2}.
When increasing the prior's strength, the posterior estimate moves towards the prior smoothly, as expected (see \cref{imperf-test:fig:ntot}).
As discussed previously, the \cref{E-ATACCC} estimates are reliable for the first 2--3 weeks, notably including the median time.
The median using $r\inform$ matched \cref{E-ATACCC}'s median estimate and is a principled choice because it is based directly on the prior work \textcite{birrellRTM2}.
Therefore, I use this estimate in the application to the CIS data.
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/CIS_ntot}
  \caption[Sensitivity of $\ntot$'s posterior to its prior.]{How the posterior estimate of $\ntot$ changes with the value of $r$ in the prior on $\ntot$.}
  \label{imperf-test:fig:ntot}
\end{figure}

>>>>

Applying the method of \cref{E-perf-test} to CIS data estimates a median duration of less than 14 days, several days shorter than the estimates from \cref{E-ATACCC} (see \cref{imperf-test:fig:problem-cis-estimates}).
The \cref{E-ATACCC} estimates are more reliable over the first 2--3 weeks due to ATACCC's frequent data collection (see \cref{E-biology-data:sec:ataccc,E-ATACCC:sec:discussion}).
This suggests an issue in the estimation using CIS data.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-imperfect-testing/CIS_perfect}}
  \caption[Estimating survival using CIS data assuming perfect testing]{Estimates of the survival function using CIS data alongside the ATACCC-based estimates of \cref{E-perf-test}. \label{imperf-test:fig:problem-cis-estimates}}
\end{figure}

@@

% Discussion and interrogation of the data led me to
I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negative results.
There are several strands of evidence supporting this hypothesis.
First, it is well known that RT-PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
Furthermore, in \cref{imperf-test:sec:simulate}, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives.
This reproduced the CIS data more faithfully (see \cref{imperf-test:fig:sim-single-pos}) and exhibited the same issues when estimating the duration distribution (not shown).
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-imperfect-testing/sim-single-positive-episodes}}
  \caption[Single positive episodes in CIS simulation]{%
    Proportion of detected episodes which have a single positive episode.
    y-axis is the number of simulations, out of \numprint{1000}, with the indicated proportion.
    Test sensitivity is $\psens$, with the varied test sensitivity referring to the model proposed in \cref{imperf-test:eq:variable-test-sensitivity}.
    The CIS data are shown as a vertical line.
  }
  \label{imperf-test:fig:sim-single-pos}
\end{figure}

@@

The qualitative increase in long episodes is robust to the choice of prior for $\ntot$, the assumed value for $\psens$ (see \cref{imperf-test:fig:cis-sensitivity}), and the choice of prior for the hazards, $\lambda_t$ (see \cref{E-perf-test:sec:parameters-priors}).
However, the survival proportion over the first 4 weeks is sensitive to these choices.
The estimate using a test sensitivity of 0.8 and $\NBc(\mu\inform, r\inform)$ give a median survival time most similar to that of \cref{E-ATACCC}.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-imperfect-testing/CIS_final}}
  \caption[Comparing estimates of the duration distribution.]{
    Duration estimates using the model in this chapter (CIS) alongside those from \cref{E-ATACCC} (ATACCC).
    Uses the informative prior on $\ntot$, the model combination prior for the hazards, and $\psensi = 0.8$.
  }
  \label{imperf-test:fig:cis-estimates}
\end{figure}
\begin{figure}
  \thisfloatpagestyle{empty}
  \makebox[\textwidth][c]{\includegraphics{cis-imperfect-testing/CIS_vary}}
  \caption[Sensitivity of CIS estimates to the prior.]{%
    (A-C) Sensitivity of CIS estimates to the value of $r$, the dispersion of the prior for $\ntot$ (a larger $r$ means a more informative prior) when $\psens = 0.8$.
    (D-F) Sensitivity of CIS estimates to the value of $\psens$ when $r = r\inform$.
    A and D: median survival time, in comparison to the ATACCC-based estimate of \cref{E-ATACCC} (shown in red).
    B and E: survival at day 50, $S_{\vec{\theta}}(50)$.
    C and F: full survival curves out to day 100.
  }
  \label{imperf-test:fig:cis-sensitivity}
  \todo[inline]{Optional improvement: Could highlight the "preferred" model which exists in both rows and is the one used subsequently, and/or include ATACCC in the far-right hand column}
\end{figure}

The estimates are sensitive to the choice of $r$, the strength of the prior on $\ntot$.
A low value for $r$, giving a very weak, almost uninformative, prior on $\ntot$ causes its posterior estimate to be much higher than the estimate from \textcite{birrellRTM2}.
When increasing the prior's strength, the posterior estimate moves towards the prior smoothly, as expected (see \cref{imperf-test:fig:ntot}).
As discussed previously, the \cref{E-ATACCC} estimates are reliable for the first 2--3 weeks, notably including the median time.
The median using $r\inform$ matched \cref{E-ATACCC}'s median estimate and is a principled choice because it is based directly on the prior work \textcite{birrellRTM2}.
Therefore, I use this estimate in the application to the CIS data.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-imperfect-testing/CIS_ntot}}
  \caption[Sensitivity of $\ntot$'s posterior to its prior.]{How the posterior estimate of $\ntot$ changes with the value of $r$ in the prior on $\ntot$.}
  \label{imperf-test:fig:ntot}
\end{figure}
